---
title: "Bias App"
# author: Rolf Grönwold, Tom Palmer, Kate Tilling
output:
  flexdashboard::flex_dashboard:
    source_code: https://github.com/remlapmot/bias-app
    vertical_layout: fill # scroll
runtime: shiny
---

```{r setup, include=FALSE}
if (!requireNamespace("DiagrammeR", quietly = TRUE)) install.packages("DiagrammeR")
library(DiagrammeR)

if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
library(tidyverse)

if (!requireNamespace("plotly", quietly = TRUE)) install.packages("plotly")
library(plotly)
```

```{r, include=FALSE}
# functions for the biases
confounderbias <-
  function(x = 0.1,
           ratio = 1,
           varc = 1,
           sigma2x = 1,
           rho = NULL,
           vary = NULL,
           varx = NULL) {
    # x is bcx
    if (is.null(rho)) {
      bcy <- x / ratio
      (x * bcy * varc) / (x ^ 2 * varc + sigma2x)
    } else {
      (rho * sqrt(vary)) / sqrt(varx)
    }
  }

colliderbias <-
  function(x = 0.1,
           ratio = 1,
           bxy = 0,
           varx = 1,
           sigma2y = 1,
           sigma2c = 1) {
    # x is bxc
    byc <- x / ratio
    vary <- bxy ^ 2 * varx + sigma2y
    varc <-
      (x ^ 2 * varx) + (byc ^ 2 * vary) + (2 * x * bxy * byc * varx) + sigma2c
    num <- -1 * (bxy * bxy + x) * (vary - bxy ^ 2 * varx) * byc
    den <- varc - varx * (bxy * byc + x) ^ 2
    num / den
  }

selectionbias <-
  function(x = 1,
           bxy = 1,
           sigma2y = 1,
           sigma2s = 1,
           varx = 1) {
    # x is bys
    bys <- x
    vary <- bxy ^ 2 * varx + sigma2y
    vars <- bys ^ 2 * vary + sigma2s
    num <- bxy * (varx * vars - bys ^ 2 * vary * varx) ^ 2
    den <- varx * vars - (bxy * bys * varx) ^ 2
    (num / den) - bxy
  }
```

# Confounding bias {data-orientation=rows}

Row {data-height=650}
---

### App 

```{r}
# ui ----
  ui = fluidPage(
    withMathJax(),
      # fluidRow(
      # column(6, grVizOutput("confdiagram", width = "250px"))
      # ),
  fluidRow(column(12, plotlyOutput("plot1"))),
    fluidRow(
      column(2,
             numericInput("ratio", "Ratio: \\(\\beta_{cx} / \\beta_{cy}\\)", value = 1, step = 0.1, width = "100px")),
      column(2,
             numericInput("varc", "\\(\\text{var}(C)\\)", value = 1, step = 0.1, width = "100px", min = 1e-6)),
      column(2,
             numericInput("sigma2x", "\\(\\sigma^2_{X}\\)", value = 1, step = 0.1, width = "100px", min = 1e-6)),
      column(2,
             numericInput("xmin", "X-axis min", value = -0.2, step = 0.1, width = "100px")),
      column(2,
             numericInput("xmax", "X-axis max", value = 0.2, step = 0.1, width = "100px"))
    )
  )
  
# server ----
  server = function(input, output) {

    # output$confdiagram <- renderGrViz({
    #   grViz({"
    #   digraph dag {
    #   graph [rankdir=TB]
    #   node [shape=plaintext, width=0.3, height=0.3]
    #   X 
    #   C 
    #   Y
    #   node [shape=circle, height=0.3, fixedsize=true]
    #   Ex [label='&epsilon;@_{x}']
    #   Ey [label='&epsilon;@_{y}']
    #   { rank = same; X Y }
    #   X -> Y [minlen=3, label='&beta;@_{xy}']
    #   C -> X [label='&beta;@_{cx}']
    #   C -> Y [minlen=.5, label='&beta;@_{cy}']
    #   C -> C [dir='both', headport='n', tailport='n']
    #   Ex -> X [label='1']
    #   Ey -> Y [label='1']
    #   }
    #   "})
    # })
    
    output$plot1 <- renderPlotly({
        ggplotly(
        ggplot(data = data.frame(x = 0), 
               mapping = aes(x = x)) + 
        stat_function(fun = confounderbias, 
                      colour = "blue",
                      args = list(varc = input$varc,
                                  sigma2x = input$sigma2x,
                                  ratio = input$ratio)) +
        scale_x_continuous(limits = c(input$xmin, input$xmax)) + 
        ylab("Bias") + 
        xlab("$\\beta_{cx}$")
      )
    })
  }

# Define as an app ----
shinyApp(ui = ui, server = server)
```

Row {data-height=350}
---

### Description

In this model, shown in the path diagram on the right, we consider the effect of omitting a confounder, \(C\), from the regression of \(Y\) on \(X\).

The bias in the estimate of \(\beta_{xy}\) when we do not adjust for \(C\) is given by: $$\text{Bias} = \frac{\beta_{cx}\beta_{cy}\text{var}(C)}{\beta_{cx}^2\text{var}(C) + \sigma_{X}^2}$$

The plot visualises this bias. Change the settings to explore how the bias changes.

### Model

```{r}
DiagrammeR::grViz("
      digraph dag {
      graph [rankdir=TB]
      node [shape=plaintext, width=0.3, height=0.3]
      X 
      C 
      Y
      node [shape=circle, height=0.3, fixedsize=true]
      Ex [label='&epsilon;@_{x}']
      Ey [label='&epsilon;@_{y}']
      { rank = same; X Y }
      X -> Y [minlen=3, label='&beta;@_{xy}']
      C -> X [label='&beta;@_{cx}']
      C -> Y [minlen=.5, label='&beta;@_{cy}']
      C -> C [dir='both', headport='n', tailport='n']
      Ex -> X [label='1']
      Ey -> Y [label='1']
      }
      ")
```

# Collider bias

```{r}
# ui ----
  ui = fluidPage(
    withMathJax(),
      fluidRow(
        column(6,
      p(
        "In this model, shown in the path diagram on the right, we consider the effect of including a collider, \\(C\\), in the regression of \\(Y\\) on \\(X\\) and \\(C\\)."),
        br(),
        p("The bias in the estimate of \\(\\beta_{xy}\\) when we adjust for \\(C\\) is given by: $$\\text{Bias} = -\\frac{(\\beta_{xy}\\beta_{yc} + \\beta_{xc})(\\text{var}(Y)- \\beta_{xy}^2\\text{var}(X))\\beta_{yc}}{\\text{var}(C) - \\text{var}(X)(\\beta_{xy}\\beta_{yc} + \\beta_{xc})^2}$$"),
        br(),
        p("The plot below visualises this bias. Change the settings to explore how the bias changes."
      )),
      column(6, grVizOutput("confdiagram", width = "250px"))
      ),
  fluidRow(column(12, plotlyOutput("plot1"))),
    fluidRow(
      column(2,
             numericInput("ratio", "Ratio: \\(\\beta_{xc} / \\beta_{yc}\\)", value = 1, step = 0.1, width = "100px")),
      column(2,
             numericInput("bxy", "\\(\\beta_{xy}\\)", value = 0, step = 0.1, width = "100px")),
      column(2,
             numericInput("varx", "\\(\\text{var}(X)\\)", value = 1, step = 0.1, width = "100px", min = 1e-6)),
      column(1,
             numericInput("sigma2y", "\\(\\sigma^2_{y}\\)", value = 1, step = 0.1, width = "100px", min = 1e-6)),
      column(1,
             numericInput("sigma2c", "\\(\\sigma^2_{c}\\)", value = 1, step = 0.1, width = "100px", min = 1e-6)),
      column(2,
             numericInput("xmin", "X-axis min", value = -0.2, step = 0.1, width = "100px")),
      column(2,
             numericInput("xmax", "X-axis max", value = 0.2, step = 0.1, width = "100px"))
    )
  )
  
# server ----
  server = function(input, output) {

    output$confdiagram <- renderGrViz({
      grViz({"
      digraph dag {
      graph [rankdir=TB, layout=neato]
      node [shape=box, height=0.4, width=0.4]
      X [pos='0,1!']
      C [pos='1,0!']
      Y [pos='2,1!']
      node [shape=circle, height=0.4, fixedsize=true]
      Ec [label='&epsilon;@_{C}', pos='1,-1!']
      Ey [label='&epsilon;@_{y}', pos='2,2!']
      { rank = same; X Y }
      X -> C [label='&beta;@_{xc}']
      Y -> C [minlen=.5, label='&beta;@_{yc}']
      X -> Y [label='&beta;@_{xy}']
      Ec -> C [label=1]
      X -> X [dir='both', headport='w']
      Ey -> Y [label=1]
      }
      "})
    })
    
    output$plot1 <- renderPlotly({
        ggplotly(
        ggplot(data = data.frame(x = 0), 
               mapping = aes(x = x)) + 
        stat_function(fun = colliderbias, 
                      colour = "blue",
                      args = list(sigma2c = input$sigma2c,
                                  sigma2y = input$sigma2y,
                                  ratio = input$ratio,
                                  bxy = input$bxy,
                                  varx = input$varx)) +
        scale_x_continuous(limits = c(input$xmin, input$xmax)) + 
        ylab("Bias") + 
        xlab("$\\beta_{xc}$")
      )
    })
  }

# Define as an app ----
shinyApp(ui = ui, server = server)
```

# Selection bias

```{r}
# ui ----
  ui = fluidPage(
    withMathJax(),
      fluidRow(
        column(6,
      p(
        "In this model, shown in the path diagram on the right, we consider the effect of adjusting for a variable, S, which is caused by the outcome, i.e. we fit the regression of Y on X and S."),
        br(),
        p("The plot below visualises this bias. Change the settings to explore how the bias changes."
      )),
      column(6, grVizOutput("seldiagram", width = "250px"))
      ),
  fluidRow(column(12, plotlyOutput("plot1"))),
    fluidRow(
      column(2,
             numericInput("bxy", "\\(\\beta_{xy}\\)", value = 1, step = 0.1, width = "100px")),
      column(2,
             numericInput("sigma2y", "\\(\\sigma^2_{Y}\\)", value = 1, step = 0.1, width = "100px", min = 1e-6)),
      column(2,
             numericInput("sigma2s", "\\(\\sigma^2_{S}\\)", value = 1, step = 0.1, width = "100px", min = 1e-6)),
      column(2,
             numericInput("varx", "\\(\\text{var}(X)\\)", value = 1, step = 0.1, width = "100px", min = 1e-6)),
      column(2,
             numericInput("xmin", "X-axis min", value = -0.2, step = 0.1, width = "100px")),
      column(2,
             numericInput("xmax", "X-axis max", value = 0.2, step = 0.1, width = "100px"))
    )
  )
  
# server ----
  server = function(input, output) {

    output$seldiagram <- renderGrViz({
      grViz({"
      digraph dag {
      graph [rankdir=TB]
      node [shape=box, width=0.3, height=0.3]
      X 
      Y
      S
      { rank = same; X Y S }
      node [shape=circle, height=0.35, fixedsize=true]
      Ey [label='&epsilon;@_{Y}']
      Es [label='&epsilon;@_{S}']
      { rank = same; Ey Es }
      X -> Y [minlen=3, label='&beta;@_{xy}']
      Y -> S [minlen=3, label='&beta;@_{ys}']
      X -> X [dir='both', headport='w', tailport='w']
      Ey -> Y [label=1]
      Es -> S [label=1]
      }
      "})
    })
    
    output$plot1 <- renderPlotly({
        ggplotly(
        ggplot(data = data.frame(x = 0), 
               mapping = aes(x = x)) + 
        stat_function(fun = selectionbias, 
                      colour = "blue",
                      args = list(bxy = input$bxy,
                                  sigma2y = input$sigma2y,
                                  sigma2s = input$sigma2s,
                                  varx = input$varx)) +
        scale_x_continuous(limits = c(input$xmin, input$xmax)) + 
        ylab("Bias") + 
        xlab("$\\beta_{ys}$")
      )
    })
  }

# Define as an app ----
shinyApp(ui = ui, server = server)
```

# Figure 2 {data-orientation=rows}

<!-- Row {data-height=350} -->
<!-- --- -->

<!-- ### Description -->

<!-- Simulation results -->

Row {data-height=1000}
---------------------

### Plot

```{r}
E.OLS.unadj <- function(V) {
	s.u1=V[1]; 	s.u2=V[2]
	s.x=V[3]; 	s.m=V[4]
	s.y=V[5]; 	b.xy=V[6]
	b.xm=V[7]; 	b.u1x=V[8]
	b.u1m=V[9]; b.u1y=V[10]
	b.my=V[11]; b.u2m=V[12]
	b.u2y=V[13]
	var.x 	<- b.u1x^2*s.u1^2 + s.x^2
	cov.u1x 	<- s.u1^2*b.u1x
	cov.xy 	<- b.xy*var.x + b.xm*b.my*var.x + b.u1x*s.u1^2*(b.u1y + b.u1m*b.my)
	var.m 	<-  b.u1m^2*s.u1^2 + b.u2m^2*s.u2^2 + b.xm^2*var.x + 2*b.u1m*b.xm*cov.u1x + s.m^2 
	cov.xm 	<- b.xm*var.x + b.u1x*s.u1^2*b.u1m
	cov.my 	<- b.my*var.m + b.u1m*s.u1^2*b.u1y + b.u2m*s.u2^2*b.u2y + b.xm*b.u1x*s.u1^2*b.u1y + b.xm*var.x*b.xy + b.u1m*s.u1^2*b.u1x*b.xy

	ref 	<- b.xy + b.xm*b.my		# reference: total effect:
	unadj <- b.xy + b.xm*b.my + b.u1x*s.u1^2*(b.u1y + b.u1m*b.my)/var.x 	# without adjustment for M:
	unadj - ref
}

E.OLS.adj <- function(V) {
	s.u1=V[1]; 	s.u2=V[2]
	s.x=V[3]; 	s.m=V[4]
	s.y=V[5]; 	b.xy=V[6]
	b.xm=V[7]; 	b.u1x=V[8]
	b.u1m=V[9]; b.u1y=V[10]
	b.my=V[11]; b.u2m=V[12]
	b.u2y=V[13]
	var.x 	<- b.u1x^2*s.u1^2 + s.x^2
	cov.u1x 	<- s.u1^2*b.u1x
	cov.xy 	<- b.xy*var.x + b.xm*b.my*var.x + b.u1x*s.u1^2*(b.u1y + b.u1m*b.my)
	var.m 	<-  b.u1m^2*s.u1^2 + b.u2m^2*s.u2^2 + b.xm^2*var.x + 2*b.u1m*b.xm*cov.u1x + s.m^2 
	cov.xm 	<- b.xm*var.x + b.u1x*s.u1^2*b.u1m
	cov.my 	<- b.my*var.m + b.u1m*s.u1^2*b.u1y + b.u2m*s.u2^2*b.u2y + b.xm*b.u1x*s.u1^2*b.u1y + b.xm*var.x*b.xy + b.u1m*s.u1^2*b.u1x*b.xy

	ref 	<- b.xy + b.xm*b.my		# reference: total effect:
	adj 	<- (cov.xy*var.m - cov.my*cov.xm)/(var.x*var.m - cov.xm^2) 	# with adjustment for M:
  adj - ref
}

s.u1 <- s.u2 <- s.x <- s.m <- s.y <- 1
b.xy 	<- 0
b.xm 	<- seq(0,1,.25)
b.u1x <- seq(0,1,.25)
b.u1m <- seq(0,1,.25)
b.u1y <- seq(0,1,.25)
b.my 	<- seq(0,1,.25)
b.u2m <- seq(0,1,.25)
b.u2y	<- seq(0,1,.25)

d <- expand.grid(s.u1, s.u2, s.x, s.m, s.y,
		b.xy, b.xm, b.u1x, b.u1m, 
		b.u1y, b.my, b.u2m, b.u2y)

Mat.unadj <- apply(d,1, E.OLS.unadj)
Mat.adj <- apply(d,1, E.OLS.adj)

dat <- data.frame(d)
names(dat) <- c("s.u1", "s.u2", "s.x", "s.m", "s.y",
		"b.xy", "b.xm", "b.u1x", "b.u1m", 
		"b.u1y", "b.my", "b.u2m", "b.u2y")
dat2 <- rbind(dat,dat)
dat2$adjust <- rep(0:1,each=dim(dat)[1])
dat2$bias <- c(Mat.unadj,Mat.adj)

par(mfrow = c(2,4))
# layout(matrix(c(1,2,3,4,5,6,7,8), 
#              ncol=4, byrow=TRUE))
# par(mai = rep(0.5, 4))
boxplot(dat2$bias~dat2$adjust + dat2$b.xm, 
	col=c("white","grey"),las=1,xlab='', main=expression(beta[xm]), 
	ylab='bias',xaxt='n', cex.main=2)
axis(side=1,at=2*1:5-.5, labels=b.xm); abline(0,0,lty=1)

boxplot(dat2$bias~dat2$adjust + dat2$b.my, 
	col=c("white","grey"),las=1,xlab='', main=expression(beta[my]), 
	ylab='bias',xaxt='n', cex.main=2)
axis(side=1,at=2*1:5-.5, labels=b.my); abline(0,0,lty=1)

boxplot(dat2$bias~dat2$adjust + dat2$b.u1x, 
	col=c("white","grey"),las=1,xlab='', main=expression(beta[u1x]), 
	ylab='bias',xaxt='n', cex.main=2)
axis(side=1,at=2*1:5-.5, labels=b.u1x); abline(0,0,lty=1)

boxplot(dat2$bias~dat2$adjust + dat2$b.u1m, 
	col=c("white","grey"),las=1,xlab='', main=expression(beta[u1m]), 
	ylab='bias',xaxt='n', cex.main=2)
axis(side=1,at=2*1:5-.5, labels=b.u1m); abline(0,0,lty=1)

boxplot(dat2$bias~dat2$adjust + dat2$b.u1y, 
	col=c("white","grey"),las=1,xlab='', main=expression(beta[u1y]), 
	ylab='bias',xaxt='n', cex.main=2)
axis(side=1,at=2*1:5-.5, labels=b.u1y); abline(0,0,lty=1)

boxplot(dat2$bias~dat2$adjust + dat2$b.u2m, 
	col=c("white","grey"),las=1,xlab='', main=expression(beta[u2m]), 
	ylab='bias',xaxt='n', cex.main=2)
axis(side=1,at=2*1:5-.5, labels=b.u2m); abline(0,0,lty=1)

boxplot(dat2$bias~dat2$adjust + dat2$b.u2y, 
	col=c("white","grey"),las=1,xlab='', main=expression(beta[u2y]), 
	ylab='bias',xaxt='n', cex.main=2)
axis(side=1,at=2*1:5-.5, labels=b.u2y); abline(0,0,lty=1)

# par(mai = rep(0, 4))
# plot.new()
# plot(-1, -1, xlim = c(0,1),
#      ylim = c(0,1),
#      xaxt = 'n',
#      yaxt = 'n',
#      xlab = '',
#      ylab = '',
#      bty = 'n')
# legend(x = "center",
#        legend = c("unadjusted", "adjusted"),
#        fill = c("white", "grey"), cex = 1)
plot(-1,-1,xlim=c(0,1), ylim=c(0,1),xaxt='n',yaxt='n',xlab='',ylab='', bty='n')
legend(x="center",
       y="center",
       legend=c("unadjusted","adjusted"),
       fill=c("white","grey"), 
       cex=.85)
```


<!--
# Mediator with confounding
-->

# Figure 4 {data-orientation=rows}

Row {data-height=650}
---------------------

### Plot

```{r}
E.OLS.unadj <- function(V) {
	s.u1=V[1]; 	s.u2=V[2]
	s.x=V[3]; 	s.m=V[4]
	s.y=V[5]; 	b.xy=V[6]
	b.xm=V[7]; 	b.u1x=V[8]
	b.u1m=V[9]; b.u1y=V[10]
	b.my=V[11]; b.u2m=V[12]
	b.u2y=V[13]
	var.x 	<- b.u1x^2*s.u1^2 + s.x^2
	cov.u1x 	<- s.u1^2*b.u1x
	cov.xy 	<- b.xy*var.x + b.xm*b.my*var.x + b.u1x*s.u1^2*(b.u1y + b.u1m*b.my)
	var.m 	<-  b.u1m^2*s.u1^2 + b.u2m^2*s.u2^2 + b.xm^2*var.x + 2*b.u1m*b.xm*cov.u1x + s.m^2 
	cov.xm 	<- b.xm*var.x + b.u1x*s.u1^2*b.u1m
	cov.my 	<- b.my*var.m + b.u1m*s.u1^2*b.u1y + b.u2m*s.u2^2*b.u2y + b.xm*b.u1x*s.u1^2*b.u1y + b.xm*var.x*b.xy + b.u1m*s.u1^2*b.u1x*b.xy

	ref 	<- b.xy + b.xm*b.my		# reference: total effect:
	unadj <- b.xy + b.xm*b.my + b.u1x*s.u1^2*(b.u1y + b.u1m*b.my)/var.x 	# without adjustment for M:
	out 	<- unadj - ref
	return(out)}

E.OLS.adj <- function(V) {
	s.u1=V[1]; 	s.u2=V[2]
	s.x=V[3]; 	s.m=V[4]
	s.y=V[5]; 	b.xy=V[6]
	b.xm=V[7]; 	b.u1x=V[8]
	b.u1m=V[9]; b.u1y=V[10]
	b.my=V[11]; b.u2m=V[12]
	b.u2y=V[13]
	var.x 	<- b.u1x^2*s.u1^2 + s.x^2
	cov.u1x 	<- s.u1^2*b.u1x
	cov.xy 	<- b.xy*var.x + b.xm*b.my*var.x + b.u1x*s.u1^2*(b.u1y + b.u1m*b.my)
	var.m 	<-  b.u1m^2*s.u1^2 + b.u2m^2*s.u2^2 + b.xm^2*var.x + 2*b.u1m*b.xm*cov.u1x + s.m^2 
	cov.xm 	<- b.xm*var.x + b.u1x*s.u1^2*b.u1m
	cov.my 	<- b.my*var.m + b.u1m*s.u1^2*b.u1y + b.u2m*s.u2^2*b.u2y + b.xm*b.u1x*s.u1^2*b.u1y + b.xm*var.x*b.xy + b.u1m*s.u1^2*b.u1x*b.xy

  ref 	<- b.xy + b.xm*b.my		# reference: total effect:
  adj 	<- (cov.xy*var.m - cov.my*cov.xm)/(var.x*var.m - cov.xm^2) 	# with adjustment for M:
  adj - ref
}

s.u1 	<- s.u2 <- s.x <- s.m <- s.y <- 1
b.xy 	<- 0
b.xm 	<- seq(-.5, .5, .05) 	# b.xm=b.u1m
b.u1x <- c(-0.3, .1, .4)	# b.u1x = b.u1y
b.u1y <- 1; b.u1m <- 1
b.my 	<- 0; 
b.u2m <- b.u2y <- 0

d 		<- expand.grid(s.u1, s.u2, s.x, s.m, s.y, b.xy, b.xm, b.u1x, b.u1m, b.u1y, b.my, b.u2m, b.u2y)
names(d) 	<- c("s.u1", "s.u2", "s.x", "s.m", "s.y","b.xy", "b.xm", "b.u1x", "b.u1m",	"b.u1y", "b.my", "b.u2m", "b.u2y")
d["b.u1m"] <- d["b.xm"]		# b.xm=b.u1m
d["b.u1y"] <- d["b.u1x"]	# b.u1x = b.u1y

Mat.unadj 	<- apply(d,1, E.OLS.unadj)
Mat.adj 	<- apply(d,1, E.OLS.adj)

dat 		<- data.frame(d)
dat$unadj 	<- apply(d,1, E.OLS.unadj)
dat$adj 	<- apply(d,1, E.OLS.adj)

B.U1Y <- 0.1
plot(dat$b.xm[dat$b.u1y==B.U1Y], dat$unadj[dat$b.u1y==B.U1Y], type='l', ylim=c(-.05,.25), las=1, lty=3,
		ylab='bias', xlab=expression(paste(beta[xm]," , ", beta[u1m])))
lines(dat$b.xm[dat$b.u1y==B.U1Y], dat$adj[dat$b.u1y==B.U1Y], lty=2)
text(x=0,y=-.015, expression(paste(beta[u1x],", ",beta[u1y]," = ",0.1)))

B.U1Y <- 0.4
lines(dat$b.xm[dat$b.u1y==B.U1Y], dat$unadj[dat$b.u1y==B.U1Y], lty=3)
lines(dat$b.xm[dat$b.u1y==B.U1Y], dat$adj[dat$b.u1y==B.U1Y], lty=2)
text(x=0,y=.15,expression(paste(beta[u1x],", ",beta[u1y]," = ",0.4)))

B.U1Y = -0.3
lines(dat$b.xm[dat$b.u1y==B.U1Y], dat$unadj[dat$b.u1y==B.U1Y], lty=3)
lines(dat$b.xm[dat$b.u1y==B.U1Y], dat$adj[dat$b.u1y==B.U1Y], lty=2)
text(x=0,y=0.07,expression(paste(beta[u1x],", ",beta[u1y]," = ",-0.3)))

legend('topleft', 
       legend = c("unadjusted", "adjusted"), 
       lty = 3:2, 
       bty = 'n')
```

# Figure 6 {data-orientation=rows}

Row {data-height=650}
---------------------

### Plot

```{r}
# function that quantifies bias due to unmeasured confounding:
UC <- function(b.mu, b.mx, b.xu, b.ym, b.yu) b.xu*b.yu / (b.xu^2 + 1)

# function that quantifies bias when conditioning on an intermediate:
CIUC <- function(b.mu, b.mx, b.xu, b.ym, b.yu) {
	(b.xu*b.yu - b.mu*b.mx*b.yu)/(b.xu^2 + b.mu^2 + 1) - b.mx*b.ym
}

# function that quantifies difference in absolute bias:
delta.bias <- function(b.mu, b.mx, b.xu, b.ym, b.yu) {
	abs(UC(b.mu,b.mx,b.xu,b.ym,b.yu)) - abs(CIUC(b.mu,b.mx,b.xu,b.ym,b.yu))
}

# ranges of values for different parameters:
b_mu <- seq(0.9, 1.1, length.out = 10)
b_mx <- seq(0.05, 0.15, length.out = 10)
b_xu <- seq(0.15, 0.25, length.out = 10)
b_ym <- seq(1.0, 1.2, length.out = 10)
b_yu <- seq(0.6, 0.8, length.out = 10)

input <- expand.grid(b_mu, b_mx, b_xu, b_ym, b_yu)
diff.bias <- numeric(dim(input)[1])

for (i in 1:length(diff.bias)){
	diff.bias[i] <- delta.bias(input[i,1], input[i,2], input[i,3], input[i,4], input[i,5])
}
# mean(diff.bias > 0)

plot(density(diff.bias), 
	yaxt = 'n', 
	xlab = 'Difference in absolute bias', 
	main = '',
	xlim = c(-.3,.3),
	ylab = '')
	lines(x = c(0,0), y = c(0,1e2), lty = 2)
	text(x = -.15, y = 5.5,'Favours not conditioning', cex = .8)
	text(x = .2, y = 5.5, 'Favours conditioning', cex = .8)
```

Row 
---

### Description

This is for Example 2 in the preprint.

Difference in absolute bias plot to compare bias under 2 analysis strategies:

1. Analysis without adjustment (bias due to unmeasured confounding): UC
2. Analysis with adjustment for an intermediate variable (proxy for unmeasured confounding): CIUC

# About

This app accompanies our paper *Conditioning on a mediator*. The preprint is available [here](https://osf.io/sj7ch/).

Authors:

- Rolf Grönwold, Department of Clinical Epidemiology/Department of Medical Statistics, Leiden University Medical Centre, Leiden, Netherlands
- Tom Palmer, MRC Integrative Epidemiology Unit and Population Health Sciences, Bristol Medical School, University of Bristol, Bristol, UK
- Kate Tilling, MRC Integrative Epidemiology Unit and Population Health Sciences, Bristol Medical School, University of Bristol, Bristol, UK

Please email, <tom.palmer@bristol.ac.uk>, with any questions or comments or create an issue in the GitHub repo [here](https://github.com/remlapmot/bias-app/issues).